\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    urlcolor=blue
}

\lstset{
    backgroundcolor=\color{gray!10},
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    rulecolor=\color{gray!40},
    xleftmargin=6pt,
    xrightmargin=6pt,
    aboveskip=10pt,
    belowskip=10pt,
    columns=fullflexible,
    keepspaces=true
}

\title{Semantic Segmentation Layer: Latency Investigation\\[4pt]
\large IGVC\_ROS2 perception pipeline --- root cause analysis and bench-test plan}
\author{Multi-agent research synthesis (Phase 1--3)}
\date{2026-04-26}

\begin{document}
\maketitle

\section{Executive summary}

The ``delay'' observed with the semantic segmentation layer is two unrelated
problems sharing a name. They must be diagnosed and fixed separately.

\begin{longtable}{p{0.22\linewidth} p{0.22\linewidth} p{0.27\linewidth} p{0.22\linewidth}}
\toprule
\textbf{Symptom} & \textbf{Magnitude} & \textbf{Root cause} & \textbf{Fix} \\
\midrule
Costmap ``comet trail'' of stale \texttt{LETHAL} cells trailing behind the
robot &
$5\,\mathrm{s} \times 1\,\mathrm{m/s} = 5\,\mathrm{m}$ &
Kiwicampus C++ data race: \texttt{updateBounds} iterates
\texttt{temporal\_tile\_map\_} without holding its mutex; competing with
\texttt{bufferSegmentation}. Workaround forced \texttt{tile\_map\_decay\_time}
$1.5 \to 5.0\,\mathrm{s}$. &
4-line C++ patch. Confirmed safe (consistent lock order, no deadlock). Same
bug present on jazzy branch. \\
\midrule
Per-frame perception latency, single ZED X &
$\sim$100 ms end-to-end; \textbf{94\,\% GPU} on one camera &
\emph{Unknown.} Stereolabs' published spec is \textbf{11\,\% GPU} at
HD1080@30 NEURAL\_LIGHT. The 8$\times$ overshoot at half the documented
frame rate is anomalous --- top suspects: \texttt{point\_cloud\_res:
COMPACT} doing CUDA work, CUDA-OpenCV \texttt{cv2.resize} in
\texttt{perception\_node}, or wrapper/SDK version drift. &
Cannot diagnose without Jetson access. 8-step on-Jetson checklist
isolates the cause in $<$10 min. \\
\bottomrule
\end{longtable}

The first fix is a clean software change ready to ship. The second
\emph{cannot} be confirmed from this machine --- it requires
\texttt{nvidia-smi dmon} while toggling settings.

\section{Background}

\subsection{Hardware}

\begin{itemize}[leftmargin=2em,nosep]
\item Jetson Orin AGX 64 GB (the 64 GB SKU; full 275 TOPS, 2048 CUDA cores Ampere)
\item ZED Link Quad GMSL2 capture card
\item 3$\times$ ZED X cameras (front: serial 49910017; left: 43779087; right: TBD)
\item ZED SDK 5.2.0, \texttt{zed-ros2-wrapper} v5.2.2
\item L4T R36 (JetPack 6), ROS 2 Humble
\end{itemize}

\subsection{Software in the perception path}

\begin{enumerate}[leftmargin=2em,nosep]
\item \texttt{zed\_wrapper} composable component publishes
  \texttt{rgb/color/rect/image} (540$\times$960 after \texttt{pub\_downscale\_factor: 2.0})
  and \texttt{point\_cloud/cloud\_registered} (256$\times$448 organized,
  \texttt{point\_cloud\_res: COMPACT}).
\item \texttt{avros\_perception/perception\_node.py} \texttt{ApproximateTimeSynchronizer}
  pairs them, resizes RGB to cloud H$\times$W, runs the HSV pipeline,
  republishes mask + confidence + cloud + overlay (all carrying the image's
  \texttt{header.stamp}).
\item Kiwicampus \texttt{semantic\_segmentation\_layer} (Nav2 plugin)
  consumes mask + cloud + \texttt{LabelInfo}, projects mask pixels into 3D
  via the cloud, accumulates into \texttt{temporal\_tile\_map\_}, writes
  cost into the local costmap on \texttt{updateBounds}/\texttt{updateCosts}.
\end{enumerate}

\subsection{Empirical numbers from CHANGELOG\_2026-04-25}

\begin{itemize}[leftmargin=2em,nosep]
\item Image, cloud, and perception mask all within 100--120 ms of \texttt{header.stamp} (mean 102 ms)
\item Topic rates: \texttt{rgb/color/rect/image} 15.0 Hz, \texttt{semantic\_mask} 15.0 Hz, \texttt{semantic\_points} 15.0 Hz
\item Tegrastats: GPU at 94\,\% utilization (single camera, HD1080@15, NEURAL\_LIGHT)
\item Master costmap with \texttt{tile\_map\_decay\_time: 5.0}: 290 stable LETHAL cells when tape in view; 0 within 5 s of covering camera
\end{itemize}

\section{Problem 1: the kiwicampus mutex bug}

\subsection{Diagnosis}

The 2026-04-25 changelog \S 3 documents the bug; Phase 1 of this investigation
verified it directly against the source under
\texttt{src/semantic\_segmentation\_layer/}.

\paragraph{Two separate recursive mutexes:}
\begin{itemize}[leftmargin=2em,nosep]
\item \texttt{SegmentationBuffer::lock\_} --- \texttt{segmentation\_buffer.hpp:786}
\item \texttt{SegmentationTileMap::lock\_} --- \texttt{segmentation\_buffer.hpp:347}
\end{itemize}

Both \texttt{std::recursive\_mutex}, completely independent.

\paragraph{The race:}
\begin{itemize}[leftmargin=2em,nosep]
\item \texttt{bufferSegmentation} (15 Hz cloud-sync callback) acquires
  \texttt{temporal\_tile\_map\_->lock()} to mutate the map
  (\texttt{segmentation\_buffer.cpp:203})
\item \texttt{updateBounds} (5 Hz costmap update loop) acquires only
  \texttt{buffer->lock()} and iterates the tile\_map without holding its
  own mutex (\texttt{semantic\_segmentation\_layer.cpp:298--331})
\item \texttt{updateBounds} also calls
  \texttt{purgeOldObservations(node->now())} --- node clock --- while
  observations were stamped with \texttt{cloud.header.stamp}
  (\texttt{segmentation\_buffer.cpp:141})
\end{itemize}

\paragraph{Symptom (observed at runtime):}
DEBUG-level \texttt{nav2\_costmap\_2d} logging revealed that 90\,\% of update
cycles passed degenerate bounds to \texttt{updateCosts}:

\begin{lstlisting}
48 cycles: x: [249, 1] y: [249, 1]    <-- min > max -> silent early exit
 3 cycles: x: [120, 158] y: [99, 144]  <-- valid, includes our cells
 1 cycle:  x: [0, 250] y: [0, 250]     <-- initial reset
\end{lstlisting}

\paragraph{Workaround already in production:}
\texttt{tile\_map\_decay\_time: 1.5 $\to$ 5.0\,s} in
\texttt{nav2\_params\_humble.yaml} \S\texttt{semantic\_layer.\{front,left,right\}}.
Buffer stays populated long enough that the for-loop sees real tiles. Cost:
5 s of stale lethal cells trail behind a moving robot.

\subsection{The patch}

Lock order (buffer first, tile\_map second) is consistent across all call
sites, so the patch is deadlock-safe.

\begin{lstlisting}
--- a/src/semantic_segmentation_layer.cpp
+++ b/src/semantic_segmentation_layer.cpp
@@ -298,6 +298,7 @@ void SemanticSegmentationLayer::updateBounds(...)
   for (auto& tile_map_pair : segmentation_tile_maps)
   {
     auto buffer = tile_map_pair.second;
     buffer->lock();
+    tile_map_pair.first->lock();

     tile_map_pair.first->purgeOldObservations(current_time);
@@ -328,6 +329,7 @@
       touch(tile_world_coords.x, tile_world_coords.y, min_x, min_y, max_x, max_y);
     }
+    tile_map_pair.first->unlock();
     buffer->unlock();
   }
\end{lstlisting}

\subsection{Delivery strategy (recommended)}

\textbf{Strategy (a):} keep upstream \texttt{humble} branch pin in
\texttt{avros.repos}, add a second patch file alongside the existing PR \#1
patch.

\begin{enumerate}[leftmargin=2em,nosep]
\item Save diff as \texttt{src/avros\_bringup/patches/kiwicampus\_pr2\_mutex.patch}
\item Extend \texttt{scripts/apply\_kiwicampus\_patches.sh} to apply both
  patches idempotently
\item \texttt{colcon build --packages-select semantic\_segmentation\_layer}
\item After verification, drop \texttt{tile\_map\_decay\_time: 5.0 $\to$ 2.0} in
  all three \texttt{semantic\_layer} source blocks
\item Open PR upstream at
  \url{https://github.com/kiwicampus/semantic_segmentation_layer} citing the
  \S 3 analysis from CHANGELOG\_2026-04-25
\end{enumerate}

\section{Problem 2: 100 ms latency / 94\,\% GPU anomaly}

\subsection{Diagnosis}

Phase 1 found a contradiction:

\begin{itemize}[leftmargin=2em,nosep]
\item Code-reading of \texttt{perception\_node.py} budgets the in-node work at
  \textbf{33--54 ms} (resize 8--12 ms; HSV blur 6--12 ms; cvtColor 1--2 ms;
  three \texttt{inRange} 1.5--3 ms; morphology 2--4 ms; overlay loop 2--4 ms;
  \texttt{cv\_bridge} round-trips 6--9 ms).
\item Stereolabs' published table for HD1080@30 NEURAL\_LIGHT on AGX Orin
  is \textbf{11\,\% GPU}.
\item Empirical: \textbf{94\,\% GPU} on \emph{one} camera at HD1080@\emph{15}
  --- 8$\times$ the documented load at half the frame rate.
\end{itemize}

Something else is loading the GPU. Top suspects, ranked:

\paragraph{1. \texttt{point\_cloud\_res: COMPACT} triggers a CUDA pass.}
The Stereolabs docs describe COMPACT as ``optimizes processing and bandwidth''
without specifying the implementation. Inference: COMPACT downsamples the
depth map and projects to XYZ via CUDA. Adding a per-frame CUDA pass on top
of NEURAL\_LIGHT inference would explain the gap.

\paragraph{2. CUDA-OpenCV in \texttt{perception\_node}.}
\texttt{perception\_node.py:288--291} unconditionally calls
\texttt{cv2.resize(bgr, (cloud.width, cloud.height), interpolation=INTER\_AREA)}
because cloud H$\times$W (256$\times$448) differs from image H$\times$W
(540$\times$960). Standard JetPack OpenCV is CPU-only, but a custom build
with \texttt{WITH\_CUDA=ON} would route this to the GPU. Confirm with
\texttt{cv2.getBuildInformation()}.

\paragraph{3. Wrapper/SDK version drift.}
\texttt{zed-ros2-wrapper} v5.2.2 must be linked against an SDK 5.2.x
matching the major.minor pin. Drift causes silent slow-path fallback. The
project's \texttt{CLAUDE.md} states SDK 5.2.0 is installed; confirm the
header reports the same.

\paragraph{Lower probability:} Foxglove bridge subscribers triggering NVENC
streaming; \texttt{grab\_compute\_capping\_fps} unset; tegrastats summing
unrelated CUDA work.

\subsection{What the perception node looks like (for reference)}

\begin{lstlisting}
# perception_node.py _on_synced (paraphrased)
bgr = cv_bridge.imgmsg_to_cv2(image, 'bgr8')          # 3-5 ms
if bgr.shape[:2] != (cloud.height, cloud.width):
    bgr = cv2.resize(bgr, (W, H), INTER_AREA)         # 8-12 ms (or GPU?)
result = self._pipeline.run(bgr)                      # 20-30 ms (HSV path)
# 4 publishes: mask, confidence, overlay, cloud-relay
\end{lstlisting}

The node itself is \emph{innocent} of the headline 100 ms. The 50--70 ms
unaccounted for must live in the wrapper or a CUDA path we haven't
identified.

\section{3-camera scaling decision}

\subsection{Configuration trade-off}

\begin{longtable}{p{0.06\linewidth} p{0.34\linewidth} p{0.20\linewidth} p{0.30\linewidth}}
\toprule
\textbf{ID} & \textbf{Config} & \textbf{Predicted GPU (Stereolabs spec)} & \textbf{Trade-off} \\
\midrule
C1 & 3$\times$ HD1080@15 NEURAL\_LIGHT (today's plan) & $\sim$33\,\% & Best detection; risky given the 94\,\% anomaly is unexplained \\
C2 & Front HD1080 + sides SVGA, all NEURAL\_LIGHT@15 & $\sim$15--20\,\% & \textbf{Recommended.} Sides keep depth $\to$ all 3 feed semantic\_layer; lateral lane visibility preserved during turns \\
C3 & Front HD1080 + sides RGB-only (\texttt{depth\_mode: NONE}) & $\sim$11\,\% & Cheap, but sides cannot feed semantic\_layer (no cloud to project mask into) \\
C4 & 3$\times$ SVGA NEURAL\_LIGHT & $\sim$10\,\% & Max headroom, sacrifices front detection range for thin lanes \\
C5 & Front NEURAL + sides NEURAL\_LIGHT, all HD1080@15 & $\sim$50\,\% & Better front depth quality, no headroom \\
\bottomrule
\end{longtable}

\subsection{Recommended: C2}

YAML edits required:

\begin{lstlisting}
# zed_left.yaml AND zed_right.yaml
/**:
  ros__parameters:
    general:
      grab_resolution: 'SVGA'      # was 'HD1080'
      grab_frame_rate: 15
    depth:
      depth_mode: 'NEURAL_LIGHT'
      max_depth: 8.0               # was 15.0; SVGA range
\end{lstlisting}

\texttt{zed\_front.yaml} unchanged.

\subsection{Composable container}

After C2 is verified, refactor the three ZED nodes into a single
\texttt{ComposableNodeContainer} with \texttt{use\_intra\_process\_comms:
true}, following Stereolabs' \texttt{zed\_ipc.launch.py} pattern. Saves
$\sim$63 MB/s of cloud serialization across 3 cameras. Mandatory before
sustained 3-camera operation.

\textbf{Driver requirement:} ZED X driver $\geq$ 1.4.1 (Stereolabs forum
thread 10853 documents an I$^2$C contention bug fixed in 1.4.0+ that breaks
multi-camera auto-exposure).

\section{Bench-test plan}

\subsection{Block A: Diagnose the 94\,\% GPU anomaly (highest leverage; $\sim$10 min)}

Run sequentially. Stop and report as soon as one gives a clear answer.

\begin{enumerate}[leftmargin=2em,nosep]
\item Confirm SDK version matches wrapper:
\begin{lstlisting}
cat /usr/local/zed/include/sl/Camera.hpp | \
  grep -E "ZED_SDK_(MAJOR|MINOR)_VERSION"
\end{lstlisting}
Expected: \texttt{MAJOR=5}, \texttt{MINOR=2}.

\item Launch sensors, confirm depth\_mode actually applied:
\begin{lstlisting}
ros2 launch avros_bringup sensors.launch.py enable_zed_front:=true
ros2 param get /zed_front/zed_node depth.depth_mode
\end{lstlisting}
Expected: \texttt{NEURAL\_LIGHT}.

\item Check whether OpenCV is CUDA-built (would route \texttt{cv2.resize} to GPU):
\begin{lstlisting}
python3 -c "import cv2; print(cv2.getBuildInformation())" | \
  grep -i cuda
\end{lstlisting}
If \texttt{CUDA=YES}, the perception node's resize is on the GPU.

\item Isolate the cloud pipeline. Edit \texttt{zed\_front.yaml} to set
  \texttt{point\_cloud\_freq: 0.0}, relaunch, then:
\begin{lstlisting}
nvidia-smi dmon -i 0 -s pucm -c 30
\end{lstlisting}
If GPU drops below $\sim$20\,\%, the cloud pipeline (probably \texttt{COMPACT}
projection) is the cost. Restore \texttt{point\_cloud\_freq: 15.0} after.

\item Toggle Foxglove bridge to rule out NVENC encoding:
\begin{lstlisting}
nvidia-smi dmon -i 0 -s pucm -c 30 &
pkill -f foxglove_bridge
sleep 10
ros2 run foxglove_bridge foxglove_bridge
\end{lstlisting}
Look for a step change in GPU \%.

\end{enumerate}

\subsection{Block B: Apply and verify the kiwicampus mutex patch ($\sim$30 min)}

\begin{enumerate}[leftmargin=2em,nosep]
\item Save the diff (Section 3.2) to \texttt{src/avros\_bringup/patches/kiwicampus\_pr2\_mutex.patch}
\item Extend \texttt{scripts/apply\_kiwicampus\_patches.sh} to apply both
  patches idempotently (use \texttt{git apply --reverse --check} guard for each)
\item Rebuild:
\begin{lstlisting}
cd ~/IGVC
rm -rf build/semantic_segmentation_layer install/semantic_segmentation_layer
colcon build --packages-select semantic_segmentation_layer --symlink-install
source install/setup.bash
\end{lstlisting}
\item With \texttt{tile\_map\_decay\_time: 5.0} (current), launch full
  navigation stack, place white tape in view, then cover the camera. Confirm
  cells fade in 5 s as before --- patch shouldn't regress.
\item Drop \texttt{tile\_map\_decay\_time: 5.0 $\to$ 2.0} in all three
  \texttt{semantic\_layer} source blocks of \texttt{nav2\_params\_humble.yaml}.
  Relaunch.
\item \textbf{Verification A} --- with tape in view, watch
  \texttt{/local\_costmap/costmap} lethal-cell count for 60 s. Expect stable
  count, no flicker, no zero-bins.
\item \textbf{Verification B} --- cover the camera, expect cells to fade in
  $\sim$2 s.
\item \textbf{Verification C} --- restart \texttt{controller\_server} with
  \texttt{--log-level nav2\_costmap\_2d:=DEBUG} and confirm bounds passed to
  \texttt{updateCosts} are valid every cycle (no more degenerate
  \texttt{x: [249, 1]} lines).
\item If all three verifications pass: commit and update CHANGELOG. If
  cell-count flickers at decay\_time=2.0, retry at 3.0 s.
\end{enumerate}

\subsection{Block C: 3-camera bring-up ($\sim$20 min, gated on A and B)}

\begin{enumerate}[leftmargin=2em,nosep]
\item Baseline single-camera GPU during driving:
\begin{lstlisting}
ros2 launch avros_bringup webui.launch.py &
ros2 launch avros_bringup sensors.launch.py enable_zed_front:=true &
tegrastats --interval 1000 > /tmp/gpu_1cam.txt &
# Drive WebUI joystick for 60 s, then:
kill %3
awk '/GR3D/ {print $4}' /tmp/gpu_1cam.txt | sort -n | tail -5
\end{lstlisting}
\item Plug in left + right ZED Xs. Verify with \texttt{ZED\_Explorer} that
  serials enumerate.
\item Edit \texttt{zed\_left.yaml} and \texttt{zed\_right.yaml} per Section 4.2.
\item Wire \texttt{enable\_zed\_left:=true enable\_zed\_right:=true} into
  \texttt{sensors.launch.py} (currently only front is plumbed --- this is
  Phase 5 launch work, see TODO list in CLAUDE.md).
\item Repeat tegrastats capture with all three cameras up. Target: max GPU
  $<$ 80\,\% for 60 s of driving. If $>$ 80\,\%, drop front to SVGA or revisit
  Block A's cause.
\end{enumerate}

\subsection{Block D: Composable container migration (gated on C)}

Refactor sensors launch to follow Stereolabs'
\texttt{zed\_ipc.launch.py} pattern: one
\texttt{ComposableNodeContainer} loading three \texttt{stereolabs::ZedCamera}
components plus the three \texttt{PerceptionNode} components, all with
\texttt{use\_intra\_process\_comms: true}. \emph{Do not combine this change
with the resolution change in Block C} --- one variable at a time.

\section{Open questions (still unanswerable from desk)}

\begin{itemize}[leftmargin=2em,nosep]
\item Whether the 94\,\% GPU number was during a representative run
  (Foxglove + RViz + perception all up) or in isolation. Block A step 1
  baselines this.
\item Whether the wrapper on Jetson is actually linked against SDK 5.2.x
  (vs 5.1 or 5.3 drift). Block A step 1 covers this.
\item Whether \texttt{tile\_map\_decay\_time: 2.0} works without false
  negatives during real driving. Block B verifications A/B/C cover this.
\item Real \texttt{zed\_front} mount offset in URDF (currently measured
  values from team sketch in CHANGELOG\_2026-04-25 \S 2). Out of scope
  for this investigation.
\end{itemize}

\section{Reference --- file locations}

\begin{longtable}{p{0.45\linewidth} p{0.50\linewidth}}
\toprule
\textbf{File} & \textbf{Purpose} \\
\midrule
\texttt{src/semantic\_segmentation\_layer/src/semantic\_segmentation\_layer.cpp} & Where the patch goes (around line 298--331) \\
\texttt{src/semantic\_segmentation\_layer/src/segmentation\_buffer.cpp} & Documents the timestamp source for purge logic (line 141) \\
\texttt{src/avros\_bringup/patches/kiwicampus\_pr1.patch} & Existing Humble build-fix patch (PR \#1) \\
\texttt{scripts/apply\_kiwicampus\_patches.sh} & Idempotent patch runner; needs extending for PR \#2 \\
\texttt{src/avros\_bringup/config/nav2\_params\_humble.yaml} & \texttt{semantic\_layer} block (lines 188--251); decay\_time lives here \\
\texttt{src/avros\_bringup/config/zed\_front.yaml} & ZED front overrides (HD1080, NEURAL\_LIGHT, COMPACT) \\
\texttt{src/avros\_bringup/config/zed\_left.yaml}, \texttt{zed\_right.yaml} & Edit per Section 4.2 for C2 \\
\texttt{src/avros\_perception/avros\_perception/perception\_node.py} & The bridge node; resize at lines 288--291 \\
\texttt{src/avros\_bringup/launch/sensors.launch.py} & Where left/right launch args need plumbing for Phase 5 \\
\texttt{docs/CHANGELOG\_2026-04-25.md} & Source of truth for the bug discovery and workaround \\
\bottomrule
\end{longtable}

\section{External references}

\begin{itemize}[leftmargin=2em,nosep]
\item Stereolabs Depth Modes: \url{https://www.stereolabs.com/docs/depth-sensing/depth-modes}
\item Stereolabs ROS2 Composition: \url{https://www.stereolabs.com/docs/ros2/ros2-composition}
\item \texttt{zed-ros2-examples} multi-camera tutorial: \url{https://github.com/stereolabs/zed-ros2-examples/blob/master/tutorials/zed_multi_camera/README.md}
\item Stereolabs forum --- multi-camera scaling on AGX Orin: \url{https://community.stereolabs.com/t/performance-on-agx-orin/4406}
\item Stereolabs forum --- ZED Link Quad I$^2$C fix in driver 1.4.0: \url{https://community.stereolabs.com/t/62-frame-drops-with-auto-exposure-enabled-on-4x-zed-x-gmsl2-data-attached/10853}
\item Kiwicampus repo: \url{https://github.com/kiwicampus/semantic_segmentation_layer}
\item Upstream Nav2 segmentation reference: \url{https://github.com/pepisg/nav2_segmentation_demo}
\end{itemize}

\end{document}
